\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.

\usepackage[hyphens]{url}
\usepackage{hyperref} 
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{etoolbox}
\newbool{colored}
\IfFileExists{./colored}{\booltrue{colored}}{\boolfalse{colored}}
\newbool{ascii}
\IfFileExists{./ascii}{\booltrue{ascii}}{\boolfalse{ascii}}

\usepackage{subcaption}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{numprint}
\usepackage{multirow}
\usepackage{xcolor}
%\usepackage{todonotes}
% \setuptodonotes{inline}
\newcommand{\todo}[1]{{\bf TODO: #1}}

\usepackage{tikz}
\usetikzlibrary{positioning,fit,arrows.meta,backgrounds}

% (Gabor) I usually write with frenchspacing on but turned it off now.
% Caveat: With nonfrenchspacing, comments in algorithm2e use double-spacing after ':', '.', etc. which looks bad.
% \frenchspacing

% (Scott) I usually use the minted package for code blocks configured as follows
% You put code inside \begin{cplus} \end{cplus}
\usepackage{minted}
\setminted{fontsize=\footnotesize,baselinestretch=.97,linenos,frame=lines,xleftmargin=6pt,numbersep=3pt,mathescape=true,escapeinside=||,bgcolor=bg}
\usemintedstyle{default}
\definecolor{bg}{rgb}{0.97,0.97,0.97}
\newminted{cpp}{}
\newenvironment{cplus}{\VerbatimEnvironment\begin{cppcode}}{\end{cppcode}}
\newmintinline[cplusinl,mathescape]{cpp}{}



% TODO: remove in submitted version
\thispagestyle{plain}\pagestyle{plain}

\input{algorithm-commands}

\input{commands}
\input{graphblas-commands}
\input{listings}

\begin{document}

%\title{LAGraph: A Graph Algorithm and Network Analysis Library for GraphBLAS}
\title{LAGraph: Linear Algebra, Network Analysis Libraries, and the Study of Graph Algorithms}
\author{ % Tim D modified the ordering: Gabor, then alphabetical
\IEEEauthorblockN{
    G\'abor Sz\'arnyas\IEEEauthorrefmark{5},
    David A. Bader\IEEEauthorrefmark{6},
    Timothy A. Davis\IEEEauthorrefmark{1},
    James Kitchen\IEEEauthorrefmark{3}, \\
    Timothy G. Mattson\IEEEauthorrefmark{2},
    Scott McMillan\IEEEauthorrefmark{4},
    Erik Welch\IEEEauthorrefmark{3}
} \\
\IEEEauthorblockA{
    \IEEEauthorrefmark{1}Texas A\&M University
}
\IEEEauthorblockA{
    \IEEEauthorrefmark{2}Parallel Computing Labs, Intel Corporation, Ocean Park, WA
}
\IEEEauthorblockA{
    \IEEEauthorrefmark{3}Anaconda, Inc.
}
\IEEEauthorblockA{
    \IEEEauthorrefmark{4}
    Software Engineering Institute, Carnegie Mellon University, Pittsburgh, PA
}
\IEEEauthorblockA{
    \IEEEauthorrefmark{5}CWI Amsterdam, The Netherlands
}
\IEEEauthorblockA{
    \IEEEauthorrefmark{6}New Jersey Institute of Technology
}
}

\maketitle

\begin{abstract}

Graphs and graph algorithms can be expressed in terms of 
linear algebra. The GraphBLAS are a library of low level building blocks for such algorithms.
The GraphBLAS target algorithm \emph{developers}.  The LAGraph project
targets graph algorithm \emph{users} with high-level algorithms common in network
analysis.   In this paper, we describe 
the first release of the LAGraph library.  We describe the design decisions behind the
library, the contents of the library, and performance data using the GAP benchmark suite.
LAGraph, however, is much more than a library development project.  It is also a
project to document and analyze the full range of algorithms enabled by the GraphBLAS.  To that end
we have developed a compact (and hopefully intuitive) notation for describing
these algorithms.  In this paper, we present that 
notation with examples from the GAP benchmark suite.  

\end{abstract}

\begin{IEEEkeywords}
Graph Processing, Graph Algorithms, Graph Analytics, Linear Algebra, GraphBLAS
\end{IEEEkeywords}

\input{intro}

%TimM removed this figure.  We are short on space and this doesn't add enough to justify the space it consumes.
%\input{fig-architecture}

\input{design-decisions}

\input{notation}

\input{algorithms}

\input{utility}

\input{evaluation}

\input{conclusion}

\input{acknowledgements}

%\clearpage
\bibliographystyle{IEEEtranS}
\bibliography{ms}

\end{document}

% TODOs for the camera-ready
% - make sure page numbering is turned off
% - check whether GraphBLAS v2.0 is out (maybe cross-cite other GrAPL paper)
% - check whether cited CoRR papers (PIUMA, GraphBLAST, FPGA survey, ...) have been accepted
